---
title: "Emergency Admission Analysis"
author: "Ansuman Jaiswal"
date: "2026-03-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Question 1 First, read in the data set and save the contents to a dataframe df. What are the variable types of each of the columns?(2 mark)

```{r}
df <- read.csv("EmergencyAdmission_Dirty.csv")
str(df)
```

MatchIPM, Patient_Num, AdmissionAge, Death, Operative, Complications and Endoscopic are numeric (integer) variables.

Dates, gender and clinical info (Outcome, DiagnosisText, etc.) are stored as character variables.

##Question 2 Do a summary statistics description of the data using R with summary and head.(3 marks)

```{r}
# view some row
head(df)
# summary statistics
summary(df)
```

The head(df) output shows some first few observation of the data set,allow us to understand how the data is structured.it display different variables such as MatchIPM, Patient_num, MHAdmissionDate,Discharge and so on.

The summary(df) output shows whole summary of the data set, allows us to understand how the data is structured like it give us minmum, maximum,mean, median, length, quartiles and help us to understand the distribution of data.its also help to understand that the data contain both catagorical and numeric.

Overall, the dataset appears to have a mix of different variable types, and the summary statistics provide a quick overview of the distribution and characteristics of the data.

## Question 3 The hospital has reverted there is some ‘missing data’ in the columns, but is unable to tell you where the missing data is or the columns. Write R code to determine where the missing data and how many of them. What will you do with them? (There is no ‘model’ answer here, but you have to justify your reasons of what to do to the missing data.)

```{r}
# count missing values of each colum
colSums(is.na(df))
# total number of missing values in dataset
sum(is.na(df))
# Display rows with missing values
df[!complete.cases(df), ]
```

To handle the missing data, different approaches can be taken depending on the situation. If the number of missing values is small, it is reasonable to remove those rows using na.omit(df) so that they do not affect the analysis.There are 14 missing values in total, all in the Death column. I would not impute them because Death is an important outcome variable. Since 14 out of 15,297 rows is very small, I would either (a) keep these rows in the dataset but exclude Death when it is NA from analyses where it is the outcome, or (b) remove these 14 rows using df_clean <- na.omit(df) and clearly report this.

```{r}
# Remove missing values
df_clean <- na.omit(df)
```

However, if the missing values are significant or the data is important, it is better to replace them instead of removing them. For example, missing values in numeric columns can be replaced using the mean to retain the overall dataset information.

## Question 4 Write code to determine the proportion of admission that results in an operational procedure.

```{r}
# Check how many admissions have an operation (non-missing OperationDate)
operation_count <- sum(!is.na(df$OperationDate))

# Total number of admissions
total_count <- nrow(df)

# Proportion
proportion <- operation_count / total_count

# Display result
proportion
```

## Question 5 The hospital also wants to know what the average lengths of stay are for the admission resulting from the Emergency department. This may be obtained by subtracting the columns MHAdmissionDate and the DischargeDate. Write the R output to determine this average length of days by creating a new variable ‘LOS’. (Hint: You will need to do transformation of the data to a numeric format as well.) 

```{r}
# convert columns to data format
df$MHAdmissionDate <- as.Date(df$MHAdmissionDate, format ="%d-%b-%y")
df$DischargeDate <- as.Date(df$DischargeDate, format ="%d-%b-%y")

# create length of stay 
df$LOS <- as.numeric(df$DischargeDate - df$MHAdmissionDate)
# VIEW LOS
head(df$LOS)
# calculate average LOS 
mean_LOS <- mean(df$LOS , na.rm = TRUE)
# display results 
mean_LOS
```


## Question 6	The hospital is of the opinion that this LOS is not accurate in determining the average length of stay.  There are stays that may be especially long and distort the average. There are two ways to remedy this. 
a.	Compute the mode and median of this LOS instead and revert to the hospital. 
b.	Filter off the stays that are longer than 30 days from the calculation. (Note: in your filtering, you must still keep a copy of the original data rows and not delete them. 

## (a) Compute the median and mode of LOS

```{r}
# Median of LOS
median_LOS <- median(df$LOS, na.rm = TRUE)
median_LOS

# Mode of LOS (custom function)
mode_function <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

mode_LOS <- mode_function(df$LOS)
mode_LOS
```
Median is robust to very long stays, so it gives a better central value than the mean when there are outliers. Mode shows the most common length of stay.

## (B)reate a filtered dataset (without deleting original data)
```{r}
df_filtered <- df[df$LOS <= 30, ]

# Calculate new average LOS
mean_LOS_filtered <- mean(df_filtered$LOS, na.rm = TRUE)
mean_LOS_filtered
```
By removing stays longer than 30 days from the calculation, we reduce the influence of rare, very long admissions and get an average more representative of typical patients.
The filtered mean is lower than the original mean, confirming that very long stays were pulling the average up.