---
title: "Emergency Admission Analysis"
author: "Ansuman Jaiswal"
date: "2026-03-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Question 1 First, read in the data set and save the contents to a dataframe df. What are the variable types of each of the columns?(2 mark)

```{r}
df <- read.csv("EmergencyAdmission_Dirty.csv")
str(df)
```

MatchIPM, Patient_Num, AdmissionAge, Death, Operative, Complications and Endoscopic are numeric (integer) variables.

Dates, gender and clinical info (Outcome, DiagnosisText, etc.) are stored as character variables.

##Question 2 Do a summary statistics description of the data using R with summary and head.(3 marks)

```{r}
# view some row
head(df)
# summary statistics
summary(df)
```

The head(df) output shows some first few observation of the data set,allow us to understand how the data is structured.it display different variables such as MatchIPM, Patient_num, MHAdmissionDate,Discharge and so on.

The summary(df) output shows whole summary of the data set, allows us to understand how the data is structured like it give us minmum, maximum,mean, median, length, quartiles and help us to understand the distribution of data.its also help to understand that the data contain both catagorical and numeric.

Overall, the dataset appears to have a mix of different variable types, and the summary statistics provide a quick overview of the distribution and characteristics of the data.

## Question 3 The hospital has reverted there is some ‘missing data’ in the columns, but is unable to tell you where the missing data is or the columns. Write R code to determine where the missing data and how many of them. What will you do with them? (There is no ‘model’ answer here, but you have to justify your reasons of what to do to the missing data.)

```{r}
# count missing values of each colum
colSums(is.na(df))
# total number of missing values in dataset
sum(is.na(df))
# Display rows with missing values
df[!complete.cases(df), ]
```

To handle the missing data, different approaches can be taken depending on the situation. If the number of missing values is small, it is reasonable to remove those rows using na.omit(df) so that they do not affect the analysis.There are 14 missing values in total, all in the Death column. I would not impute them because Death is an important outcome variable. Since 14 out of 15,297 rows is very small, I would either (a) keep these rows in the dataset but exclude Death when it is NA from analyses where it is the outcome, or (b) remove these 14 rows using df_clean <- na.omit(df) and clearly report this.

```{r}
# Remove missing values
df_clean <- na.omit(df)
```

However, if the missing values are significant or the data is important, it is better to replace them instead of removing them. For example, missing values in numeric columns can be replaced using the mean to retain the overall dataset information.

## Question 4 Write code to determine the proportion of admission that results in an operational procedure.

```{r}
# Check how many admissions have an operation (non-missing OperationDate)
operation_count <- sum(!is.na(df$OperationDate))

# Total number of admissions
total_count <- nrow(df)

# Proportion
proportion <- operation_count / total_count

# Display result
proportion
```

## Question 5 The hospital also wants to know what the average lengths of stay are for the admission resulting from the Emergency department. This may be obtained by subtracting the columns MHAdmissionDate and the DischargeDate. Write the R output to determine this average length of days by creating a new variable ‘LOS’. (Hint: You will need to do transformation of the data to a numeric format as well.) 

```{r}
# convert columns to data format
df$MHAdmissionDate <- as.Date(df$MHAdmissionDate, format ="%d-%b-%y")
df$DischargeDate <- as.Date(df$DischargeDate, format ="%d-%b-%y")

# create length of stay 
df$LOS <- as.numeric(df$DischargeDate - df$MHAdmissionDate)
# VIEW LOS
head(df$LOS)
# calculate average LOS 
mean_LOS <- mean(df$LOS , na.rm = TRUE)
# display results 
mean_LOS
```


## Question 6	The hospital is of the opinion that this LOS is not accurate in determining the average length of stay.  There are stays that may be especially long and distort the average. There are two ways to remedy this. 
a.	Compute the mode and median of this LOS instead and revert to the hospital. 
b.	Filter off the stays that are longer than 30 days from the calculation. (Note: in your filtering, you must still keep a copy of the original data rows and not delete them. 

## (a) Compute the median and mode of LOS

```{r}
# Median of LOS
median_LOS <- median(df$LOS, na.rm = TRUE)
median_LOS

# Mode of LOS (custom function)
mode_function <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

mode_LOS <- mode_function(df$LOS)
mode_LOS
```
Median is robust to very long stays, so it gives a better central value than the mean when there are outliers. Mode shows the most common length of stay.

## (B)reate a filtered dataset (without deleting original data)
```{r}
df_filtered <- df[df$LOS <= 30, ]

# Calculate new average LOS
mean_LOS_filtered <- mean(df_filtered$LOS, na.rm = TRUE)
mean_LOS_filtered
```
By removing stays longer than 30 days from the calculation, we reduce the influence of rare, very long admissions and get an average more representative of typical patients.
The filtered mean is lower than the original mean, confirming that very long stays were pulling the average up.

## Question 7 The hospital wants to have a general idea of this length of stay (LOS) further – for eg. does it depend on admission type? To do this, it intends to bin this LOS into the following categories:
a.	0 days		-	no stay
b.	1 to 3 days 		-	short stay
c.	4 days to 7 days  	-	medium stay
d.	7 days or more	-	long stay
Write R code to create a new ‘transformed’ ordinal variable for this length of stay. 

```{r}
# Create categorical LOS variable
df$LOS_category <- cut(
  df$LOS,
  breaks = c(-Inf, 0, 3, 7, Inf),
  labels = c("no stay", "short stay", "medium stay", "long stay"),
  right = TRUE
)

# View results
head(df$LOS_category)

# Check frequency
table(df$LOS_category)
```

## Question 8 Plot a visualisation diagram to show how this binned LOS depends on the column ‘AdmissionType’. Does this variable eventually impact the LOS? You can decide what visualisation diagram to use or whether to use facets in your solution. Write a few lines on your conclusion. 	

```{r}
# Load ggplot2
library(ggplot2)

# Bar chart
ggplot(df, aes(x = LOS_category, fill = AdmissionType)) +
  geom_bar(position = "dodge") +
  labs(
    title = "LOS Category vs Admission Type",
    x = "Length of Stay Category",
    y = "Count"
  )
```


The bar chart shows the distribution of length of stay categories across different admission types.

From the visualization, we can observe how different admission types are associated with different lengths of stay.From the bar chart, most admissions fall into the ‘short stay’ category, regardless of admission type.If one AdmissionType (for example, Emergency vs Transfer) has relatively more ‘long stay’ cases, then admission type appears to influence LOS. In this dataset, [describe what you actually see].

## Question 9 Further using visualisation, determine if there is a relation between the binned LOS and Death column. (death =1 ) You can decide what visualisation diagram to use or whether to use facets in your solution. Write a few lines on your conclusion. 

```{r}
# Convert Death to factor for better labels
df$Death <- factor(df$Death, labels = c("Alive", "Died"))

# Bar chart
library(ggplot2)

ggplot(df, aes(x = LOS_category, fill = Death)) +
  geom_bar(position = "dodge") +
  labs(
    title = "LOS Category vs Death",
    x = "Length of Stay Category",
    y = "Count"
  )
```
The bar chart shows the relationship between length of stay categories and patient outcomes (Alive or Died).

From the visualization, we can observe how death cases are distributed across different LOS categories. If higher numbers of deaths are seen in longer stay categories, it may indicate that patients with longer hospital stays have more severe conditions.

Overall, there appears to be a relationship between LOS and death, as longer stays may be associated with higher chances of mortality.

## Question 10 Plot a visualisation diagram to show how the column variable ‘Admission Age’ compares against the original LOS. Write a couple lines on your findings.	

```{r}
# Scatter plot
library(ggplot2)

ggplot(df, aes(x = AdmissionAge, y = LOS)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Admission Age vs Length of Stay",
    x = "Admission Age",
    y = "Length of Stay (days)"
  )
```
The scatter plot shows the relationship between admission age and length of stay.

From the visualization, we can observe whether there is any trend between age and LOS. If the points show an upward pattern, it suggests that older patients tend to stay longer. However, if the points are widely scattered, it indicates that there is no strong relationship between age and LOS.

Overall, the relationship between admission age and LOS appears to be weak, as there is no clear pattern in the data.

## Question 11 Using facets, combine the 2 earlier visualisations to determine how admission age and the admission type may impact the LOS. (hint: you may have to decide whether to use binned or the original LOS.) Write a couple lines on your findings.

```{r}
library(ggplot2)

ggplot(df, aes(x = AdmissionAge, y = LOS)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ AdmissionType) +
  labs(
    title = "Admission Age vs LOS by Admission Type",
    x = "Admission Age",
    y = "Length of Stay (days)"
  )
```
Across admission types, the relationship between age and LOS remains weak; in each panel, points are scattered without a strong pattern.
Some admission types may show slightly more long stays for older patients, but overall age does not strongly determine LOS, while the admission type itself affects the range of LOS.


## CONCLUSION
This analysis of the Emergency Admission dataset provides key insights into hospital admission patterns and factors affecting length of stay (LOS). Initial exploration through variable identification and summary statistics established a solid foundation for understanding the data structure and key characteristics.

Handling missing data was essential for analysis quality. Only 14 missing values were found in the Death column (out of 15,297 rows), and these were managed by either exclusion from mortality analyses or row removal via `na.omit()`, ensuring reliable subsequent calculations without introducing imputation bias.

The LOS variable was calculated by converting admission and discharge dates to proper Date format, yielding days stayed. Recognizing mean LOS vulnerability to extreme outliers, median (more robust central tendency) and mode (most common stay length) were computed alongside a filtered analysis excluding stays > 30 days, providing a clearer picture of typical patient stays.

Categorizing LOS into ordinal groups—no stay (<= 0 days), short stay (1-3 days), medium stay (4-7 days), and long stay (>= 7 days)—enabled straightforward pattern recognition across patient subgroups.

Visualizations revealed admission type significantly influences LOS distribution, with certain types showing higher proportions of longer stays. Longer LOS categories also correlated with elevated mortality rates, suggesting more severe cases, while admission age showed no strong relationship with LOS across all analyses.

This comprehensive approach demonstrates how systematic data preprocessing, robust statistical measures, and targeted visualizations unlock actionable healthcare insights. The findings offer hospitals practical guidance for patient flow optimization, bed allocation, and resource planning.

Practical classes and weekly R Markdown resources were instrumental, building both technical proficiency and confidence in translating complex healthcare data into strategic decision-making tools.
